Semantics for the $ \lambda $-calculus

### 2024-03-05 report/chapters/category-theory.tex
Write about the significance of local/relative cartesian closedness

\section{Dependent products and sums}\label{sec:dependent-products}

@@

However, note that the conversion from an external to an internal representation in $ \SET $ already contained a dependent sum, which is no coincidence. It turns out that in practice, we will never have a hard time obtaining dependent sums:
\begin{lemma}
  Let $ \alpha : C(B, A) $ be a morphism in a category. If the pullback functor $ \alpha^*: (C \downarrow A) \to (C \downarrow B) $ exists, it has a left adjoint given by postcomposition with $ \alpha $.
\end{lemma}
\begin{proof}
  For morphisms $ f: X \to A $, $ g: Y \to B $, the universal property of the pullback, with the following diagram
  \begin{center}
    \begin{tikzcd}
      Y \arrow[r, "\varphi"'] \arrow[rr, "\psi", bend left] \arrow[dr, "g"'] & \alpha^* X \arrow[d, "\alpha^* f"'] \arrow[r] \arrow[rd, phantom, "\lrcorner", very near start] & X \arrow[d, "f"]\\
      & B \arrow[r, "\alpha"'] & A
    \end{tikzcd}
  \end{center}
  gives an equivalence between morphisms $ \varphi: Y \to \alpha^* X $ that commute with $ g $ and $ \alpha^* f $, and morphisms $ \psi: Y \to X $ that commute with $ g $, $ f $ and $ \alpha $. In other words:
  \[ (C \downarrow A)(g \cdot \alpha, f) \simeq (C \downarrow B)(g, \alpha^*(f)), \]
  which shows the adjunction.
\end{proof}

@@

>>>>

\section{Dependent products and sums}\label{sec:dependent-products}
The following is based loosely on Section 4.1 of \cite{taylor}.

@@

However, note that the conversion from an external to an internal representation in $ \SET $ already contained a dependent sum, which is no coincidence. It turns out that in practice, we will never have a hard time obtaining dependent sums:
\begin{lemma}
  Let $ \alpha : C(B, A) $ be a morphism in a category. If the pullback functor $ \alpha^*: (C \downarrow A) \to (C \downarrow B) $ exists, it has a left adjoint given by postcomposition with $ \alpha $.
\end{lemma}
\begin{proof}
  For morphisms $ f: X \to A $, $ g: Y \to B $, the universal property of the pullback, with the following diagram
  \begin{center}
    \begin{tikzcd}
      Y \arrow[r, dashed, "\varphi"'] \arrow[rr, "\psi", bend left] \arrow[dr, "g"'] & \alpha^* X \arrow[d, "\alpha^* f"'] \arrow[r] \arrow[rd, phantom, "\lrcorner", very near start] & X \arrow[d, "f"]\\
      & B \arrow[r, "\alpha"'] & A
    \end{tikzcd}
  \end{center}
  gives an equivalence between morphisms $ \varphi: Y \to \alpha^* X $ that commute with $ g $ and $ \alpha^* f $, and morphisms $ \psi: Y \to X $ that commute with $ g $, $ f $ and $ \alpha $. In other words:
  \[ (C \downarrow A)(g \cdot \alpha, f) \simeq (C \downarrow B)(g, \alpha^*(f)), \]
  which shows the adjunction.
\end{proof}

@@

Apart from having dependent sums and products, there also is the following theorem that shows the significance locally cartesian closedness:
\begin{lemma}\label{lem:locally-cartesian-closed}
  A category $ C $ is locally cartesian closed iff $ (C \downarrow A) $ is cartesian closed for each $ A : C $.
\end{lemma}
\begin{proof}
  See the end of Section 1.3 of \cite{freyd}.
\end{proof}

### 2024-03-05 report/chapters/previous-work.tex
Write about the significance of local/relative cartesian closedness

  The category also has exponential objects
  \[ C^B = \lambda x_1, B \circ x_1 \circ C \]
  with evaluation morphism $ \epsilon_{BC}: C^B \times B \to C $ given by
  \[ \epsilon_{BC} = \lambda x_1, C(\pi_1 x_1 (B (\pi_2 x_1))), \]
  which is universal because we can lift a morphism $ f: \mathbb R(A \times B, C) $ to a morphism $ \varphi(f): \mathbb R(A, C^B) $ given by
  \[ \varphi(f) = \lambda x_1 x_2, f (x_1, x_2). \]
  Note that for $ g: \mathbb R(A, C^B) $, the inverse $ \varphi^{-1}(g) $ is given by
  \[ \epsilon \circ \langle g \circ \pi_1, B \circ \pi_2 \rangle: \mathbb R(A \times B, C). \]

@@

Now, some categories are not locally cartesian closed. That is: not all substitution functors $ \alpha^* $ exist or have a right adjoint. One way to look at this, is that not all morphisms $ X \to A $ represent an family. In these categories, we can carefully choose a subset of the morphisms to represent our indexed families. We will call a morphism representing an indexed family a \iindex{display map}. In most cases, we have quite a bit of choice which maps we want to take as our display maps. However, to make sure that indexed families are well-behaved, a class of display maps needs to have some properties:
\begin{enumerate}
  \item The pullback of a display map along any morphism exists and is a display map.
  \item The composite of two display maps is a display map.
  \item $ C $ has a terminal object and any terminal projection is a display map.
\end{enumerate}
\begin{remark}
  A maximal example of a class of display maps is, for example, in the category $ \SET $, where we can take our class of display maps to equal the class of all morphisms of $ C $.
\end{remark}
\begin{remark}
  A `minimal' example of a class of display maps is the class of (maps isomorphic to) product projections in a category with finite products. In this case, all indexed families are constant, and then dependent sums and products become binary products and exponential objects.
\end{remark}

Now, let $ C $ be a category with a class of display maps. For any $ A : C $, we get a category $ C^A $ as a full subcategory of the slice category $ C \downarrow A $, with as objects the display maps $ f: X \to A $. A morphisms from in $ f: X \to A $ to $ g: Y \to A $ is a morphism $ \varphi: X \to Y $ such that the following diagram commutes.
\begin{center}
  \begin{tikzcd}
    X \arrow[rd, "f"'] \arrow[rr, "\varphi"] & & Y \arrow[ld, "g"]\\
    & A &
  \end{tikzcd}
\end{center}
Note that for the terminal object $ I : C $, $ C^I $ is still equivalent to $ C $.

Also note that since display maps are closed under pullbacks, we can construct the pullback functor $ \alpha^*: (C \downarrow A) \to (C \downarrow B) $ for all $ \alpha: C(B, A) $.

Note that since composing two display maps gives a display map again, and since the dependent sum is given by postcomposition, the pullback functor $ \alpha^* $ has a left adjoint for all display maps $ \alpha $. That is: the fiber categories $ C^A $ have dependent sums over display maps.

The question whether the fiber categories $ C^A $ also have dependent products over display maps, brings us to the definition of relative cartesian closedness.
\begin{definition}
  A category $ C $ is \index{cartesian closed!relatively}\textit{cartesian closed relative} to a class of display maps $ D $, if the substitution functors $ \alpha^* $ along display maps have right adjoints.
\end{definition}

>>>>

  The category also has exponential objects
  \[ (C \downarrow_D B) = \lambda x_1, B \circ x_1 \circ C \]
  with evaluation morphism $ \epsilon_{BC}: (C \downarrow_D B) \times B \to C $ given by
  \[ \epsilon_{BC} = \lambda x_1, C(\pi_1 x_1 (B (\pi_2 x_1))), \]
  which is universal because we can lift a morphism $ f: \mathbb R(A \times B, C) $ to a morphism $ \varphi(f): \mathbb R(A, (C \downarrow_D B)) $ given by
  \[ \varphi(f) = \lambda x_1 x_2, f (x_1, x_2). \]
  Note that for $ g: \mathbb R(A, (C \downarrow_D B)) $, the inverse $ \varphi^{-1}(g) $ is given by
  \[ \epsilon \circ \langle g \circ \pi_1, B \circ \pi_2 \rangle: \mathbb R(A \times B, C). \]

@@

Now, some categories are not locally cartesian closed. That is: not all substitution functors $ \alpha^* $ exist or have a right adjoint. One way to look at this, is that not all morphisms $ X \to A $ represent a family of objects. In these categories, we can carefully choose a subset of the morphisms to represent our indexed families. We will call a morphism representing an indexed family a \iindex{display map}. In most cases, we have quite a bit of choice which maps we want to take as our display maps. However, to make sure that indexed families are well-behaved, a class of display maps needs to have some properties:
\begin{enumerate}
  \item The pullback of a display map along any morphism exists and is a display map.
  \item The composite of two display maps is a display map.
  \item $ C $ has a terminal object and any terminal projection is a display map.
\end{enumerate}
\begin{remark}
  A `maximal' example of a class of display maps is, for example, in the category $ \SET $, where we can take our class of display maps to equal the class of all morphisms of $ C $.
\end{remark}
\begin{remark}
  A `minimal' example of a class of display maps is the class of (maps isomorphic to) product projections in a category with finite products. In this case, all indexed families are constant, and then dependent sums and products become binary products and exponential objects.
\end{remark}

Now, let $ C $ be a category with a class of display maps $ D $. We will denote the class of display maps from $ X $ to $ A $ with $ D(X, A) \subseteq C(X, A) $. For any $ A : C $, we get a category $ (C \downarrow_D A) $ as a full subcategory of the slice category $ (C \downarrow A) $, with as objects the display maps $ f: D(X, A) $. A morphism from $ f: D(X, A) $ to $ g: D(Y, A) $ is a morphism $ \varphi: C(X, Y) $ such that the following diagram commutes.
\begin{center}
  \begin{tikzcd}
    X \arrow[rd, "f"'] \arrow[rr, "\varphi"] & & Y \arrow[ld, "g"]\\
    & A &
  \end{tikzcd}
\end{center}
Note that for the terminal object $ I : C $, $ (C \downarrow_D I) $ is still equivalent to $ C $, since every terminal projection is a display map.

Also note that since display maps are closed under pullbacks, the pullback functor $ (C \downarrow A) \to (C \downarrow B) $ for $ \alpha: C(B, A) $ restricts to a pullback functor $ \alpha^*: (C \downarrow_D A) \to (C \downarrow_D B) $.

Note that since composing two display maps gives a display map again, and since the dependent sum is given by postcomposition, $ \alpha^* $ has a left adjoint for all display maps $ \alpha $. That is: the fiber categories $ (C \downarrow_D A) $ have dependent sums over display maps.

The question whether the fiber categories $ (C \downarrow_D A) $ also have dependent products over display maps, brings us to the definition of relative cartesian closedness.
\begin{definition}
  A category $ C $ is \index{cartesian closed!relatively}\textit{cartesian closed relative} to a class of display maps $ D $, if the substitution functors $ \alpha^* $ along display maps have right adjoints.
\end{definition}

Now, analogously to \ref{lem:locally-cartesian-closed}, Taylor shows:
\begin{lemma}
  If a category $ C $ is cartesian closed relative to a class of display maps $ D $, then the fiber categories $ (C \downarrow_D A) $ are cartesian closed and the substitution functors $ \alpha^* $ preserve this structure.
\end{lemma}
\begin{proof}
  The proofs leading up to \S 4.3.7 in \cite{taylor}.
\end{proof}
